Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-10 VegOzone.tex
Change from downy birch to decideous tree

Norway spruce was experimentally found to be active already at rather low air temperatures and could reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of observed \chem{CO_2} uptake and temperature at two observation sites in Southern ($\rightarrow$ \emph{cold}) and Northern ($\rightarrow$ \emph{boreal}) Finland \citep{TB:Kolari2007}, we estimate $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume a similar acclimation for downy birch. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.\\

@@

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Tab.~\ref{tab:sensitivity_tests_gs}) and estimated as follows. Evergreen needle leaf trees show photosynthetic activity already at low temperatures \citep{TP:Wallin2013, TB:Kolari2007}. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. Net photosynthesis data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively, and is the value used for all PFTs. The resulting growing season for coniferous trees in 2019 was one month longer then in 2018. For estimating $A_\text{start}$ of downy birch, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincident with the first snow-free day at the meteorological weather station at Øvre Neiden (Sør-Varanger, NOR) which is relatively close-by.
For comparison, we would get $A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$ from the latitude model typically used for localizing $\mathrm{DO_3SE}$. By comparison, the latitude model indicates about a $2\,\unit{month}$ longer growing season. Due to a lack of quantitative field observation, we assume a latency period of about $1\,\unit{month}$ after snow-melt for perennial grassland.

@@

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The temperature acclimation dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large bias as downy birch. We also find a reversal for predicted ozone damage risk in $\mathrm{POD_1}$ between 2018 and 2019 for the boreal type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

We estimated the biomass reductions in accordance to the CLs in \citet{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature acclimation (\emph{boreal}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range in uncertainty corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of the extent of stomatal opening at low light intensities with SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass in comparison with the default generic MM parameterizations are predicted for both years. Coniferous trees are the least affected and the magnitude of total biomass reduction is independent of the choice of the parameterization within given uncertainties. This may reflect the more comprehensive research on this species or PFT feeding into the generic MM parameterization. The systematic deviation for deciduous trees is one order of magnitude larger than for coniferous trees and perennial grassland.

From the difference between 2018 and 2019, we infer a substantial interannual variability. The interannual variability is estimated to be larger than our reported systematic deviations and can be of the same order of magnitude as the effect of the temperature acclimation. This means that the choice of temperature acclimation has the potential to either revoke or amplify the risk of ozone damage for vegetation. However, the risk of ozone damage in downy birch and perennial grassland in comparison with the generic default MM parameterizations is increased, regardless.

@@

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of the growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

@@

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For coniferous trees, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care because the biomass reduction functions were not established based on subarctic species and might as well be subject to biases.

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{javis_funcs_temp_hist_birch}\\
  %\includegraphics[width=8.3cm]{javis_funcs_rad_hist_birch}
  \includegraphics[width=8.3cm]{figB3}
\caption{Construction of bespoke response functions for downy birch. (a) $f_\mathrm{T}$ and (b) $f_\mathrm{flight}$ are shown together with underlying $T_\mathrm{air}$ and $Q_0$ climatologies (probability density function - PDF), respectively. Original mapping manual parameterization is shown in comparison as solid line. Note that $Q_0$ has been truncated to $0.006$. PPFD0.8 and PPFD1.2 refer to $\alpha$ values increasing/decreasing PPFD at $f_\mathrm{light}=0.5$ by $\pm 20\,\%$, respectively.}
\label{fig:f_temp_birch}
\end{figure}

>>>>

Norway spruce was experimentally found to be active already at rather low air temperatures and could reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of observed \chem{CO_2} uptake and temperature at two observation sites in Southern ($\rightarrow$ \emph{cold}) and Northern ($\rightarrow$ \emph{boreal}) Finland \citep{TB:Kolari2007}, we estimate $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume a similar acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.\\

@@

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Tab.~\ref{tab:sensitivity_tests_gs}) and estimated as follows. Evergreen needle leaf trees show photosynthetic activity already at low temperatures \citep{TP:Wallin2013, TB:Kolari2007}. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. Net photosynthesis data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively, and is the value used for all PFTs. The resulting growing season for coniferous trees in 2019 was one month longer then in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincident with the first snow-free day at the meteorological weather station at Øvre Neiden (Sør-Varanger, NOR) which is relatively close-by.
For comparison, we would get $A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$ from the latitude model typically used for localizing $\mathrm{DO_3SE}$. By comparison, the latitude model indicates about a $2\,\unit{month}$ longer growing season. Due to a lack of quantitative field observation, we assume a latency period of about $1\,\unit{month}$ after snow-melt for perennial grassland.

@@

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Deciduous trees display the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The temperature acclimation dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large bias as deciduous trees. We also find a reversal for predicted ozone damage risk in $\mathrm{POD_1}$ between 2018 and 2019 for the boreal type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

We estimated the biomass reductions in accordance to the CLs in \citet{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature acclimation (\emph{boreal}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range in uncertainty corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of the extent of stomatal opening at low light intensities with SWP=on. For deciduous trees and perennial grassland, substantial reductions in total biomass in comparison with the default generic MM parameterizations are predicted for both years. Coniferous trees are the least affected and the magnitude of total biomass reduction is independent of the choice of the parameterization within given uncertainties. This may reflect the more comprehensive research on this species or PFT feeding into the generic MM parameterization. The systematic deviation for deciduous trees is one order of magnitude larger than for coniferous trees and perennial grassland.

From the difference between 2018 and 2019, we infer a substantial interannual variability. The interannual variability is estimated to be larger than our reported systematic deviations and can be of the same order of magnitude as the effect of the temperature acclimation. This means that the choice of temperature acclimation has the potential to either revoke or amplify the risk of ozone damage for vegetation. However, the risk of ozone damage in deciduous trees and perennial grassland in comparison with the generic MM parameterizations is increased, regardless.

@@

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for deciduous trees and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of the growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

@@

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For coniferous trees, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care because the biomass reduction functions were not established based on subarctic species and might as well be subject to biases.

@@

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{javis_funcs_temp_hist_birch}\\
  %\includegraphics[width=8.3cm]{javis_funcs_rad_hist_birch}
  \includegraphics[width=8.3cm]{figB3}
\caption{Construction of bespoke response functions for deciduous trees. (a) $f_\mathrm{T}$ and (b) $f_\mathrm{flight}$ are shown together with underlying $T_\mathrm{air}$ and $Q_0$ climatologies (probability density function - PDF), respectively. Original mapping manual parameterization is shown in comparison as solid line. Note that $Q_0$ has been truncated to $0.006$. PPFD0.8 and PPFD1.2 refer to $\alpha$ values increasing/decreasing PPFD at $f_\mathrm{light}=0.5$ by $\pm 20\,\%$, respectively.}
\label{fig:f_temp_birch}
\end{figure}
